---
title: "Causal Inference Review"
author: "Niklas Pawelzik, Justus v. Samson-Himmelstjerna"
format: html
editor: visual
---

```{r echo=FALSE}
library(dagitty)
library(ggdag)
library(ggplot2)
```

# 1. Introduction

The opioid crisis continues to pose a significant public health challenge, with a high number of overdose fatalities. A proposed strategy to address this crisis is the legalization of illicit opioids. This paper critically reviews and evaluates the study by Shane Darke and Michael Farrell titled "Would legalizing illicit opioids reduce overdose fatalities? Implications from a natural experiment," focusing on causal inference.

Darke and Farrell's study explores the impact of changes in drug purity and composition on overdose rates, aiming to assess the potential effects of legalizing heroin. However, this review highlights the need for a more rigorous causal analysis.

By examining the study's theoretical foundations, this review emphasizes the intricate causal mechanisms linking heroin legality, overdose rates, and confounding factors. It underscores the significance of considering individual and social environments, along with associated risk factors, when estimating the effects of heroin legalization on overdose fatalities.

The data used in the study primarily relies on Drug Induced Deaths (DID) and overdose cases, raising concerns about potential bias and the limited representation of the target population. Furthermore, subgroup effects are overlooked, and adequate control for confounding variables is lacking.

To address these limitations, this review suggests potential remedies, including the use of quasi-experimental methods, counterfactual scenario analysis, and the incorporation of comprehensive variables in the analysis. These approaches can enhance our understanding of the complex relationship between heroin legality and overdose fatalities, facilitating evidence-based policies to combat the opioid crisis more effectively.

In conclusion, this review underscores the importance of a robust causal analysis when investigating the potential impact of legalizing illicit opioids on overdose fatalities. By addressing the study's limitations and proposing methodological improvements, we can deepen our comprehension of the opioid crisis and inform more effective interventions to mitigate overdose rates.

# 2. Reflecting on Theoretical Foundations

## 2.1 DAG as claimed to be tested by Darke & Farrell

In the first part of the paper, the authors claim to check "the extent to which any change in legal access to heroin would affect overdose rates". According to them, it is sufficient to investigate "drug purity and drug impurity as major causes of overdose" in order to falsify this claim. Before turning to the natural experiment analysis proposed by the authors in the second half of their paper, we will take a closer look to the claimed causal mechanism of the paper. For this purpose, we visualize their model in the following DAG:

```{r echo=FALSE}
simplified_tested_relationship <- dagify(DID ~ HOD,
                                          HOD ~ P + DC,
                                          DC ~ OP, # drug components other than heroin
                                          P ~ OP, # purity/dose of heroin
                                          OP ~ L,
                                          outcome = "DID",
                                          exposure = c("P", "DC"),
                                          coords = list(x = c(L = -1,
                                                              OP = 0,
                                                              P = 1,
                                                              DC = 1,
                                                              HOD = 2,
                                                              DID = 3), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              P = 1,
                                                              DC = -1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(simplified_tested_relationship) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# P = Purity
# DC = Drug Components
# HOD = Heroin Overdose
# DID = Drug Induced Death 

```

| Symbol | Meaning            |
|--------|--------------------|
| L      | Legality           |
| OP     | Opioid Provision   |
| P      | Purity             |
| DC     | Drug Components    |
| HOD    | Heroin Overdose    |
| DID    | Drug Induced Death |

## 2.2 High Level DAGS (Unick, Rosenblum, Mars & Ciccarone)

In a paper on a related topic, Unick, et al. reviewed literature on causes for high risk of heroin overdoses. They distinguish three types of high risk variables that are subsequently added in the simplified DAG developed above:

1.  Individual Factors (IF)

2.  Risk Environment (RE)

3.  Social Environment (SE)

Looking at the DAGs below, it becomes clear that effect estimation of heroin legalization is an extremely complicated goal. Individual Factors, Social and Risk Environment behave as mediators, colliders and confounders. In the DAGs below, we included 15 additional variables all of which are well established risk factors for overdosing, we briefly discuss a selection of those relationships below.

```{r echo=FALSE}
simplified_assumed_relationship <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + IF + OP + P + DC,
                                          RE ~ IF + SE + L,
                                          IF ~ L,
                                          DC ~ OP + RE + IF, # drug components other than heroin
                                          P ~ OP + RE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6, IF = 2,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5, IF = 3,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(simplified_assumed_relationship) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 

```

| Symbol | Meaning               |
|--------|-----------------------|
| L      | Legality              |
| OP     | Opioid Provision      |
| RE     | Risk Environment      |
| SE     | Social Environment    |
| IF     | Individual Factors    |
| P      | Purity                |
| DC     | Drug Components       |
| NFNM   | Non-Fatal Near Misses |
| HOD    | Heroin Overdose       |
| DID    | Drug Induced Death    |

#### Proximal causes and individual factors

Besides polydrug use (PDU), rapid changes in heroin tolerance is among the most common individual level causes for an overdose. Reasons for such tolerance changes incarceration (I), relapse following drug treatment (R) or infrequent heroin use (IHU). All of those variables interact with legal access to heroin and its subsequent effects. For instance, legalization eases access and therefore decreases the risk for users to experience involuntary infrequent heroin use due to unavailability on the black market. In a similar vein, legal access to heroin decreases the danger of incarceration due to acquisitive crime.

```{r echo=FALSE}
assumed_relationship_IF <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          RE ~ L + SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          DC ~ OP + RE + PDU + IHU + I + R, # drug components other than heroin
                                          P ~ OP + RE + IHU + I + R, # purity/dose of heroin
                                          OP ~ L,
                                          I ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("PDU", "LOE", "I", "IHU", "R"), # used to indicate Individual Factors
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6,
                                                              I = 1, R = 2, PDU = 3, IHU = 4, LOE = 5,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5,
                                                              I= -1, R = -2, PDU = -3, IHU = -4, LOE = -5,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_IF) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# PDU = Polydrug Use
# LOE = Lack of Experience
# I = Incarceration
# IHU = Infrequent Heroin Use
# R = Relapse Following Drug Treatment

```

| Symbol | Meaning                          |
|--------|----------------------------------|
| L      | Legality                         |
| OP     | Opioid Provision                 |
| RE     | Risk Environment                 |
| SE     | Social Environment               |
| IF     | Individual Factors               |
| P      | Purity                           |
| DC     | Drug Components                  |
| NFNM   | Non-Fatal Near Misses            |
| HOD    | Heroin Overdose                  |
| DID    | Drug Induced Death               |
| PDU    | Polydrug Use                     |
| LOE    | Lack of Experience               |
| I      | Incarceration                    |
| IHU    | Infrequent Heroin Use            |
| R      | Relapse Following Drug Treatment |

#### Risk Environment

Legalization (L) has not only an effect on the risk of criminal justice risk (JCR) via potentially high-risk incarceration. Limiting legal access to heroin is also linked to "higher levels of drug law enforcement (DLE) \[that in turn\] have been linked to increased risk of overdose deaths \[16\] \[...\] increased criminal justice risk (JCR) associated with drug use leads to specific behaviors, e.g. not calling for emergency help \[(i.e. calling 911 in the United States) \[...\] following a witnessed overdose, which increase individual harm associated with drug use \[17\]." It becomes clear that "risk environment represents a number of structural components that moderate the relationship between individual factors and injection drug users' (IDUs') health" that should be considered when discussing heroin legalization. Lastly, risk environment related variables such as the access to naloxone distribution (ND) or safe injection rooms (SIR) can be mediated by legality in their effectiveness in decreasing the likelihood of heroin overdoses to be fatal.

```{r echo=FALSE}
assumed_relationship_RE <- dagify(DID ~ HOD + SE + NE + SIR + ND + JCR,
                                          NFNM ~ HOD + SE + SIR + JCR + ND,
                                          HOD ~ SE + IF + OP + P + DC + SIR + NE + MT,
                                          RE ~ L + IF + JCR,
                                          IF ~ DLE + JCR + SIR + MT + L,
                                          JCR ~ DLE + L,
                                          DLE ~ L,
                                          MT ~ L,
                                          NE ~ L,
                                          ND ~ L,
                                          DC ~ OP + SE + IF,
                                          P ~ OP + SE + IF,
                                          OP ~ L,
                                          latent = c("DLE", "JCR", "NE", "ND", "SIR", "MT"), # used to indicate risk environment
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              SE = 6, IF = 2,
                                                              DLE = 1, JCR = 2.5, NE = 3.25, ND = 4, SIR = 4.75, MT = 5.5,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              SE = 5, IF = 5,
                                                              DLE = -1.5, JCR = -3, NE = -5, ND = -7, SIR = -9, MT = -11,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 3,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_RE) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# DLE = Drug Law Enforcement 
# CJR = Criminal justice risk
# NE = Needle Exchange
# ND = Naloxone Distribution
# MT = Methadone Treatment 
# SIR = Safe Injection Rooms

```

| Symbol | Meaning               |
|--------|-----------------------|
| L      | Legality              |
| OP     | Opioid Provision      |
| RE     | Risk Environment      |
| SE     | Social Environment    |
| IF     | Individual Factors    |
| P      | Purity                |
| DC     | Drug Components       |
| NFNM   | Non-Fatal Near Misses |
| HOD    | Heroin Overdose       |
| DID    | Drug Induced Death    |
| DLE    | Drug Law Enforcement  |
| CJR    | Criminal Justice Risk |
| NE     | Needle Exchange       |
| ND     | Naloxone Distribution |
| MT     | Methadone Treatment   |
| SIR    | Safe Injection Rooms  |

#### Social Causes

Lastly, social causes need to be taken into account, as they might very well change the way the factors indicated above affect overdose and fatality risks. Being affected by poverty (EB), homelessness (HL) and other forms of social exclusion (OSE) - all listed as risk factors for drug overdose - can reinforce each other. In addition, they interact with risk environment related factors (RE) such as increased criminal justice risk or the related individual factor (IF) incarceration. In a similar vein, members of the indicated social groups are at higher risk of involuntary infrequent heroin access. All social causes indicated play a role as mediators, colliders, and/or confounders in the way legalization (L) would affect fatal overdose cases (DID).

```{r echo=FALSE}
assumed_relationship_SC <- dagify(DID ~ HOD + RE + HL + OSE,
                                          NFNM ~ HOD + RE,
                                          HOD ~ RE + IF + OP + P + DC + EB + HL + SM + OSE,
                                          RE ~ L + IF + HL + EB + SM + OSE,
                                          IF ~ L,
                                          DC ~ OP + RE + IF, 
                                          P ~ OP + RE + IF,
                                          OP ~ L,
                                          HL ~ EB + SM + OSE + IF + HOD,
                                          EB ~ SM + OSE + IF,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("EB", "HL", "SM", "OSE"),  # used to indicate social environment
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, IF = 2,
                                                              EB = 1, HL = 2.5,  SM = 3, OSE = 4,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, IF = 3,
                                                              EB = -1, HL = -2,  SM = -5, OSE = -8,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 2,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_SC) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# EB = Economic background
# HL = Homelessness
# SM = Sexual Minority Status
# OSE = Other Forms of Social Exclusion

```

| Symbol | Meaning                         |
|--------|---------------------------------|
| L      | Legality                        |
| OP     | Opioid Provision                |
| RE     | Risk Environment                |
| SE     | Social Environment              |
| IF     | Individual Factors              |
| P      | Purity                          |
| DC     | Drug Components                 |
| NFNM   | Non-Fatal Near Misses           |
| HOD    | Heroin Overdose                 |
| DID    | Drug Induced Death              |
| EB     | Economic Background             |
| HL     | Homelessness                    |
| SM     | Sexual Minority Status          |
| OSE    | Other Forms of Social Exclusion |

# 3. Reflecting on Estimants

The paper´s title claims to provide inside on the question whether "legalizing illicit opioids \[would\] reduce overdose fatalities". This would mean to estimate the total effect a change on legality had on the number of heroin induced overdose deaths. As previously pointed out, the effect of legalizing heroin on the cases of fatal overdoses involves numerous factors that mediate, confound or collide and would need careful consideration when estimating the overall effect. Looking at the dags above however, it becomes clear that at least in the first part of the paper, this is not in fact the estimant explored. Instead, the authors decided to focus on what they claim to be the two most commonly assumed drivers for decreasing fatality rates, purity and drug composition. What they investigate is the indirect effect of legalizing heroin on fatal overdoses mediated by these two variables, focusing exclusively on the arrows from Purity and Drug Components to Drug Induced Death. They furthermore limit the initially stated general claim to "provision in the manner of substances such as alcohol," thereby excluding institutional settings like supervised consumption rooms. Combining these limitations, the study therefore aims to estimate the potential impact of legalization, without additional framework measures, on changes in the purity and adulterant levels, and how these changes would affect the immediate mortality rate due to overdose. Contrary to the authors' claim, the research does not focus on the overall effect of legalizing heroin on fatal overdoses.

# 4. Reflecting on Data

The discussion above discussed a number of problematic simplifications and pointed out that the paper´s estimant is much more limited than the title suggested. In the subsequent section, we would like to turn to a different aspect, namely the data the authors based the estimation of these effects on. For the first part of the paper, the researchers quote from a number of different sources. While many of the data analyses their conclusions are beyond our reach, a common feature of these sources seems to be the focus on cases of (fatal) overdoses. Focusing only on those cases means to exclusively consider cases that, in the simplified DAGs above, have one specific value in the binary variables Heroin Overdose (HOD) and Drug Induced Death (DID). Estimating the (total, direct, or indirect) effect of legalizing heroin, however, would define the population as the population of heroin addicts. Using data primarily containing HOD or DID cases is likely to result in a sample that does not represent this target population, which needs to be taken into account.

Another issue might arise when comparing different subgroups of heroin users. When the different variables discussed interact with the effect of legalization as described above, it becomes impossible to precisely estimate the effect strength of the targeted estimant. This becomes clear when the authors aim to discuss, for instance, the effect of different levels of heroin purity on the tolerance of heroin users after periods of opioid abstinence in comparison to heroin users without abstinence. If subgroups are not considered for controlling, effect estimates might be screwed. Abstinence and its effect on tolerance are entangled in a complex way with numerous other risk factors. The authors indeed acknowledge that there "are situations in which tolerance plays a more significant role, specifically after periods of opioid abstinence". As they do not control for such variables in their limited model however, their analysis is insufficiently fine-grained to estimate the difference of effect strength; and they simply state that "such cases constitute a small minority, and it is questionable in the absence of opioid tolerance whether or not variations in purity are of relevance, as the risk of death is so high."

The issue becomes even more complicated, since being a member of a socially excluded group (OSE) means being at higher risk of an overdose, but also increases the risk and fear of Drug Law Enforcement (DLE) and incarceration. Thus, legalization can be expected to be mediated by all those variables: the behavioral change might be different across groups as legalization has a different effect on their criminal justice risk, similar effects could be expected with regard to incarceration. The point is that, in addition to the their simplified DAG, there are subsequent problematic implications for the data referenced in the article, as it becomes unclear to what extent it is fitted to control for such distinct subgroup effects.

# 5. Reflecting on the Natural Experiment

In addition to their initial discussion of purity and drug composition, the authors furthermore suggest to consider the US opioid crisis a natural experiment to estimate the effects of legalizing heroin on the number of fatal overdoses. Legal opioids would be real-world examples of widespread licit opioid provision in order to investigate how legal heroin distribution would play out. Their approach, however, seems to be mostly rely on the assumption that "the toxicological profile of these deaths \[fatal overdoses caused by licit opioids\] is consistent with that of heroin". Such a direct comparison of the overall effect across groups as seemingly implied here would only make sense if the user groups of different substances would equally composed. As soon as one assumes different user groups for legal pain killers and heroin, transferring assumptions about effects would require knowledge about the relationships of the variables within the respective DAGs.

The authors are not completely clear about the methodological considerations behind their comparisons. It seems reasonable to assume that the user group of heroin and licit opiods are not identical in terms of overdose risk factors outlined in the DAGs above. While the researchers claim comparability between one group of typical users with high risks of fatal licit opioid overdoses and typical heroin fatalities, the common trade they point out is "concomitant use". Without further investigation of the numerous additional interaction variables, such claims have to be treated with caution. These suspicions are further strengthened by the authors´ description of the second group of common fatal overdose users, namely older chronic pain patients; this group can be expected to be distinct from the typical overdose heroin user. It becomes once again clear that comparing the two populations of opioid users is hardly justifiable without further controlling.

In addition to these concerns regarding the theoretical backing and population composition, one has to carefully consider the sample composition for the natural experiment. Similar like in the paper´s first section, the researchers seem to primarily sample data from (fatal) overdoses. Thus, they only take into account cases that have a specific value for the respective binary variables. Thus, the sample at their hands is not representative of all licit opioid users but strongly biased towards users with (fatal) overdoses.

Lastly, it is worth noting that the claimed natural experiment setting does not take into account what triggers the increase in overdose fatalities Those opioids in question were legally available before there increased use, that legality is not the (only) variable that changed in the DAG. Instead, the authors state that as "prescription rates have increased, however, poisonings have increased dramatically". This puts a different focus on the crucial factors then a binary legislation variable as suggested in the paper´s title. The idea behind the natural experiment seemed to have been to simulate a counter-factual situation where a change in legalization from illegal to legal would change overdose numbers. Again, this would make it all the more crucial to ensure that all relevant variables are included in the assumed DAG in order to ensure both identification of crucial factors and transferability between subgroups.

# 6. Proposed Remedies

## 6.1 Use of Quasi-Experimental Methods

To strengthen their analysis, the authors could employ quasi-experimental methods. One such method is a difference-in-differences (DiD) approach, which compares the change in outcomes over time between a group that experienced an intervention (or policy change) and a group that did not. For instance, this approach could analyze the variation in overdose rates across states with different opioid prescription policies over time. Additionally, using instrumental variable methods could help deal with unobserved confounding variables. Alternative data sources, such as surveys or patient records, could also help offer a more representative sample of the opioid user population.

## 6.2 Counterfactual Scenario Analysis

A detailed counterfactual scenario analysis could enrich the study. One example might involve using data from jurisdictions like Portugal, where drug use has been decriminalized. By comparing overdose rates before and after this policy change and adjusting for other relevant factors, the authors could model the potential impact of similar policies in other contexts. Another counterfactual could involve synthesizing a control group via methods like propensity score matching, simulating a parallel universe in which opioid policy remained unchanged.

## 6.3 Applying a Framework for Dynamic Causal Inference

Time-varying confounders such as risk environment factors and social causes require a dynamic approach. One way is to apply Matthew Blackwell's framework for dynamic causal inference. This framework employs marginal structural models (MSMs) and inverse probability of treatment weighting (IPTW) to estimate causal effects of interventions considering time-varying confounders.

MSMs estimate causal effects in observational studies, adjusting for confounders affected by previous treatment. IPTW corrects for treatment-selection bias, creating a 'pseudo-population' independent of measured confounders. Applying this to the study involves constructing MSMs capturing dynamic relationships between heroin use, social, economic and legal status, overdose rates, and other time-varying confounders. Using IPTW, the model weighs each observation by its probability of receiving treatment given observed confounders, enabling accurate estimates of the causal effects of legalizing heroin.

# 7. Extension of Research

## 7.1 Multidimensional Approach and Robustness of Findings:
Future research should adopt a comprehensive, multidimensional approach, incorporating variables such as healthcare access, drug potency, polydrug use prevalence, and socio-economic factors, along with the legal status of heroin. Coupling this with causal models like Directed Acyclic Graphs (DAGs) will yield a more holistic understanding of heroin use and the associated risks. Sensitivity analysis should also be employed to scrutinize the robustness of the findings, ensuring that the results are reliable despite potential confounding variables.

## 7.2 Reconsidering Propensity Score Matching and Counterfactual Scenario:
While propensity score matching has been a common technique for estimating causal effects in observational studies, it's crucial to reassess its efficacy and reliability. Possible shortcomings and biases introduced by propensity score matching should be critically examined, and alternative approaches to dealing with confounding should be explored.

Furthermore, future studies could consider design-based methods such as instrumental variable methods. These methods can provide more credible causal estimates by effectively creating counterfactual scenarios, thus increasing the validity of research findings and providing a more robust basis for policy implications.

# Conclusion

In conclusion, the study by Shane Darke and Michael Farrell, titled "Would legalizing illicit opioids reduce overdose fatalities? Implications from a natural experiment," provides valuable insights into the complex relationship between heroin legality and overdose rates. However, this review highlights several areas that require further attention to strengthen the study's findings and implications.

The analysis emphasizes the limited estimant explored in the paper, focusing primarily on the indirect effects of heroin legalization mediated by purity and drug composition. The research findings suggest that changes in these variables, without additional framework measures like institutional settings, may not effectively reduce overdose rates. However, it is important to note that the study's focus is not on the overall effect of legalizing heroin on the total number of fatal overdoses.

Critically examining the data sources used, the review raises concerns about the representativeness of the sample, which primarily includes cases of Drug Induced Deaths (DID) and overdoses. This limitation implies that the sample may not accurately reflect the target population of heroin users, potentially introducing bias in the conclusions drawn.

While the authors propose the opioid crisis in the US as a natural experiment, caution is warranted in comparing the effects of legal pharmaceutical opioids with those of legalized heroin. The transferability of findings between these two substances is not adequately supported, given the potential differences in user profiles, usage patterns, and associated risk factors.

To strengthen the analysis, future research could employ quasi-experimental methods, control for confounding variables, and conduct counterfactual scenario analyses. Additionally, incorporating more comprehensive data, considering multiple dimensions of the opioid crisis, and utilizing causal models like Directed Acyclic Graphs (DAGs) would enhance our understanding of the complex interplay between heroin legality and overdose fatalities.

While the paper contributes valuable insights, further research is needed to address the identified limitations and refine our understanding of the implications of legalizing illicit opioids on reducing overdose fatalities. These endeavors will guide evidence-based policies and interventions to effectively combat the ongoing opioid crisis.
