---
title: "CI"
author: "Niklas Pawelzik"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dagitty)
library(ggdag)
library(ggplot2)
```


## High Level DAGS


```{r}
simplified_assumed_relationship <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + IF + OP + P + DC,
                                          RE ~ IF + SE,
                                          DC ~ OP + RE + IF, # drug components other than heroin
                                          P ~ OP + RE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6, IF = 2,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5, IF = 3,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(simplified_assumed_relationship) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 

```

## Proximal causes and individual factors

"The proximal causes and individual factors associated with heroin-related overdose (HOD) are
well known [1,3,6,7]. Heroin overdoses are frequently due to polydrug use (PDU), specifically when heroin is combined
with central nervous system depressants such as alcohol and benzodiazepines that contribute to respiratory failure
[8]. A lack of experience (LOE) in heroin use and rapid changes in tolerance due to incarceration (I), relapse following drug
treatment (R) or infrequent heroin use (IHU) have been linked to higher rates of overdose [6,9–11]"

```{r}
assumed_relationship_IF <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          RE ~ SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          DC ~ OP + RE + PDU + IHU + I + R, # drug components other than heroin
                                          P ~ OP + RE + IHU + I + R, # purity/dose of heroin
                                          OP ~ L,
                                          I ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("PDU", "LOE", "I", "IHU", "R"), # used to indicate Individual Factors
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6,
                                                              I = 1, R = 2, PDU = 3, IHU = 4, LOE = 5,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5,
                                                              I= -1, R = -2, PDU = -3, IHU = -4, LOE = -5,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_IF) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# PDU = Polydrug Use
# LOE = Lack of Experience
# I = Incarceration
# IHU = Infrequent Heroin Use
# R = Relapse Following Drug Treatment

```

## Social Causes

"Moving upstream, social causes such as poverty (EB), home-lessness (HL), sexual minority status (SM) or other forms of social
exclusion (OSE) have also been linked to increased risk of drug overdose [12–15]"

```{r}
assumed_relationship_SC <- dagify(DID ~ HOD + RE + HL + OSE,
                                          NFNM ~ HOD + RE,
                                          HOD ~ RE + IF + OP + P + DC + EB + HL + SM + OSE,
                                          RE ~ IF + HL + EB + SM + OSE,
                                          DC ~ OP + RE + IF, # drug components other than heroin
                                          P ~ OP + RE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          HL ~ EB + SM + OSE + IF + HOD,
                                          EB ~ SM + OSE + IF,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("EB", "HL", "SM", "OSE"),  # used to indicate social environment
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, IF = 2,
                                                              EB = 1, HL = 2.5,  SM = 3, OSE = 4,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, IF = 3,
                                                              EB = -1, HL = -2,  SM = -5, OSE = -8,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_SC) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# EB = Economic background
# HL = Homelessness
# SM = Sexual Minority Status
# OSE = Other Forms of Social Exclusion

```


## Risk environment

The risk environment represents a number of structural components that moderate the relationship
between individual factors and injection drug users’ (IDUs’) health. For example, higher levels of drug law
enforcement (DLE) have been linked to increased risk of overdose deaths [16]. The hypothesized agent of action is that
increased criminal justice risk (CJR) associated with drug use leads to specific behaviors, e.g. not calling for emergency
help (i.e. calling 911 in the United States) following a witnessed overdose, which increase individual harm
associated with drug use [17]. Other structural factors such as needle exchange (NE), naloxone distribution (ND), safe
injection rooms (SIR) and the availability of methadone treatment (MT) can reduce the risk of overdose or harm associated
with overdose in areas where these services are available

```{r}
assumed_relationship_RE <- dagify(DID ~ HOD + SE + NE + SIR + ND,
                                          NFNM ~ HOD + SE + SIR + JCR + ND,
                                          HOD ~ SE + IF + OP + P + DC + SIR + NE + MT,
                                          RE ~ IF + JCR,
                                          JCR ~ DLE,
                                          DLE ~ L,
                                          DC ~ OP + SE + IF, # drug components other than heroin
                                          P ~ OP + SE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          latent = c("DLE", "JCR", "NE", "ND", "SIR", "MT"), # used to indicate risk environment
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              SE = 6, IF = 2,
                                                              DLE = 1, JCR = 2, NE = 3, ND = 4, SIR = 5, MT = 6,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              SE = 4, IF = 3,
                                                              DLE = -1, JCR = -2, NE = -3, ND = -4, SIR = -5, MT = -6,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_RE) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# DLE = Drug Law Enforcement 
# CJR = Criminal justice risk
# NE = Needle Exchange
# ND = Naloxone Distribution
# MT = Methadone Treatment 
# SIR = Safe Injection Rooms

```

## PLease ignore below 

```{r}
# PLEASE IGNORE!!
commonly_assumed_relationship <- dagify(DID ~ HOD + CJR, # drug induced death
                                  HOD ~ P + DC + PDU + LOE + I + IHU + R +
                                        EB + HL + SM + OSE + SIR, # heroin-related overdose
                                  DC ~ OP, # drug components other than heroin
                                  P ~ OP + ND + NE, # purity/dose of heroin
                                  NFNM ~ HOD, # non-fatal near misses
                                  OP ~ L, # legalization and opioid provision
                                  CJR ~ DLE,
                        outcome = "DID",
                        # latent = c("P", "OP"),
                        exposure = "OP",
                        coords = list(x = c(L = 0,
                                            DID = 15, HOD = 13,
                                            OP = 1,  P = 2, DC = 2, NFNM = 15,
                                            IHU = 2, R = 3, PDU = 4, LOE = 5, I = 6, 
                                            EB = 2, HL = 3,  SM = 4, OSE = 5,
                                            DLE = 5, JCR = 6, NE = 7, ND = 8, SIR = 9, MT = 4),
                                      y = c(OP = 1, L = 1, DID = 1, HOD = 1, P = 1.5, DC = 0.5, NFNM = 1.5, PDU = 2, LOE = 2, I = 2, IHU = 2, R = 2,
                                            EB = 0, HL = 0,  SM = 0, OSE = 0,
                                            DLE = 1.5, JCR = 1.5, NE = 1.5, ND = 1.5, SIR = 1.5, MT = 1.5)))

ggdag_status(commonly_assumed_relationship) +
  theme_dag() +
  guides(color = "none")  # Turn off legend

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
