---
title: "Causal Inference Paper"
author: "Niklas Pawelzik, Justus v. Samson-Himmelstjerna"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dagitty)
library(ggdag)
library(ggplot2)
```

## DAG as claimed to be tested by Shane Darke & Michael Farrell

"In the community setting, where most drug use occurs, the extent to which any change in legal access to heroin would affect overdose rates depends upon the validity of the assumptions underpinning the argument: drug purity and drug impurity as major causes of overdose. We will now address each of these assumptions, before examining lessons to be drawn from the prescription opioid epidemic of recent years" 
(Would legalizing illicit opioids reduce overdose fatalities? Implications from a natural experiment)

```{r}
simplified_tested_relationship <- dagify(DID ~ HOD + E_2,
                                          HOD ~ OP + P + DC + E_1 + E_2,
                                          DC ~ OP, # drug components other than heroin
                                          P ~ OP, # purity/dose of heroin
                                          OP ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -1,
                                                              OP = 0,
                                                              P = 1,
                                                              DC = 1,
                                                              HOD = 2,
                                                              DID = 3,
                                                              E_1 = 1,
                                                              E_2 = 4), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              P = 1,
                                                              DC = -1,
                                                              HOD = 0,
                                                              DID = 0,
                                                              E_1 = 0,
                                                              E_2 = 1)))

ggdag_status(simplified_tested_relationship) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 

```

## Problems with the Data

Potential problems with the data used when investing these causal links are:

- the strong focus on DID cases:

"The absence of a strong association between purity and overdose is consistent with the demographic and toxicological characteristics of overdose cases. The typical fatal case is a long-term, dependent, daily injecting drug user (IDU) aged in their 30s or older [1–6], and not the younger, inexperienced user with low tolerance."

- having polydrug users as a maior issue might explain why a lot of heroin addicts are among the people dying from legal opioids. Thus, Using heroin and using opioids are very hard to control for when checking overdose fatalities

"Polydrug toxicity has emerged as the major factor in heroin overdoses. The overwhelming majority of opioid overdoses, both fatal and non-fatal, involve multiple CNS depressants, most notably alcohol and benzodiazepines [12–15,24–27,38,39]. Co-administration of other depressant drugs can substantially increase the likelihood of a fatal outcome, due to the combined respiratory depressant effects of these drugs."

- complicated aim to compare users after periods of opioid abstinence and non-abstinence. If not controlling correctly for these subgroups, effects might be screwed.

"There are situations in which tolerance plays a more significant role, specifically after periods of opioid abstinence, such as the period immediately following imprisonment release or detoxification. The risk of death in these specific situations is elevated substantially [15,44–49]. However, such cases constitute a small minority, and it is questionable in the absence of opioid tolerance whether or not variations in purity are of relevance, as the risk of death is so high."

## Problems with the natural Experiment

- aims to treat heroin like any other  opioid in order to be able to perceive real-world example of widespread licit opioid provision

"The situation observed in the pharmaceutical opioids epidemic is of direct relevance to our discussion, as it provides a real-world example of widespread licit opioid provision. We have widely available, labelled, legal opioids of known concentration, free of impurities. As prescription rates have increased, however, poisonings have increased dramatically. Of particular interest, a large proportion of these deaths are established heroin users who have begun to use this more readily available opioid. They are now injecting a licit opioid (albeit often not prescribed to them), rather than illicit heroin. Despite the fact that the drug is an impurity-free pharmaceutical of known dosage, toxicity deaths among this group have risen dramatically."

- first distinct group: Probably not relevant for heroin comparison, older group of chronic pain patients

- second distinct group: possibly similar to heroin group, but finding is that "the toxicology mirrors those of other opioids, including heroin, in overwhelmingly involving the concomitant use of other CNS depressants". However, we do not see the ratio of users to users that fatally overdose, we only see users that fatally overdose. Heroin users that combine opioids and other substances.

"Importantly, the toxicological profile of these deaths is consistent with that of heroin and other opioid deaths. Wider use has resulted in more deaths."

"Who is dying from these prescribed drugs? Cases comprise two distinct groups: established opioid-using IDUs, a sizable proportion of whom who inject the tablets, and a larger, older group of chronic pain patients [49,53,55,56,58,60]. Interestingly, we see again that the age profile of cases is skewed heavily towards older users, rather than young, inexperienced (and possibly low opioid-tolerant) individuals. In as many as half of cases the drug was not prescribed to the decedent [49,53,55,56]. Importantly, the toxicology mirrors those of other opioids, including heroin, in overwhelmingly involving the concomitant use of other CNS depressants, most notably benzodiazepines, alcohol and other opioids [52,53,55,56,58,60]."

- authors proposing: (re-)defining intervention: not dose and purity through legality
- increase rates due to wider availability and more widespread, unsupervised use of these drugs
- not backed by empirical findings(?)
- authors proposing: provision of injectable heroin to be consumed solely in treatment settings
- much closer to a natural experiment. intervention/controlling on several variables, but group more coherent as both are heroin consumers

"Known dose and purity do not protect and any such provision, we argue, would not reduce overdose rates, but actually increase rates due to wider availability and more widespread, unsupervised use of these drugs. A far more restricted model, which could theoretically make a major impact on overdose rates, would be the provision of injectable heroin to be consumed solely in treatment settings, such has been trialled in Switzerland and the United Kingdom [18,19]. Rates of mortality are, indeed, significantly lower in such trials, a fact shared with treatment more broadly [3]. Importantly, if an adverse reaction occurs, there is medical support present. It is arguable, however, whether this would be comprehensively provided in most jurisdictions."

## High Level DAGS (George Unick, Daniel Rosenblum, Sarah Mars & Daniel Ciccarone)

Unick, Rosenblum, Mars and Ciccarone review literature on causes for high risk of heroin overdoses. They distinguish three types of variables that should be considered and that are subsequently added in the simplified DAG developed above:

### 1. Individual Factors (IF)

### 2. Risk Environment (RE)

### 3. Social Environment (SE)

```{r}
simplified_assumed_relationship <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + IF + OP + P + DC,
                                          RE ~ IF + SE,
                                          DC ~ OP + RE + IF, # drug components other than heroin
                                          P ~ OP + RE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6, IF = 2,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5, IF = 3,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(simplified_assumed_relationship) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 

```

## Proximal causes and individual factors

"The proximal causes and individual factors associated with heroin-related overdose (HOD) are well known [1,3,6,7]. Heroin overdoses are frequently due to polydrug use (PDU), specifically when heroin is combined with central nervous system depressants such as alcohol and benzodiazepines that contribute to respiratory failure [8]. A lack of experience (LOE) in heroin use and rapid changes in tolerance due to incarceration (I), relapse following drug treatment (R) or infrequent heroin use (IHU) have been linked to higher rates of overdose [6,9–11]"
(The relationship between US heroin market dynamics and heroin-related overdose, 1992–2008)

```{r}
assumed_relationship_IF <- dagify(DID ~ HOD + RE + SE,
                                          NFNM ~ HOD + RE + SE,
                                          HOD ~ RE + SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          RE ~ SE + PDU + LOE + I + IHU + R + OP + P + DC,
                                          DC ~ OP + RE + PDU + IHU + I + R, # drug components other than heroin
                                          P ~ OP + RE + IHU + I + R, # purity/dose of heroin
                                          OP ~ L,
                                          I ~ L,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("PDU", "LOE", "I", "IHU", "R"), # used to indicate Individual Factors
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, SE = 6,
                                                              I = 1, R = 2, PDU = 3, IHU = 4, LOE = 5,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, SE = 5,
                                                              I= -1, R = -2, PDU = -3, IHU = -4, LOE = -5,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_IF) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# PDU = Polydrug Use
# LOE = Lack of Experience
# I = Incarceration
# IHU = Infrequent Heroin Use
# R = Relapse Following Drug Treatment

```

## Social Causes

"Moving upstream, social causes such as poverty (EB), home-lessness (HL), sexual minority status (SM) or other forms of social exclusion (OSE) have also been linked to increased risk of drug overdose [12–15]"
(The relationship between US heroin market dynamics and heroin-related overdose, 1992–2008)

```{r}
assumed_relationship_SC <- dagify(DID ~ HOD + RE + HL + OSE,
                                          NFNM ~ HOD + RE,
                                          HOD ~ RE + IF + OP + P + DC + EB + HL + SM + OSE,
                                          RE ~ IF + HL + EB + SM + OSE,
                                          DC ~ OP + RE + IF, # drug components other than heroin
                                          P ~ OP + RE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          HL ~ EB + SM + OSE + IF + HOD,
                                          EB ~ SM + OSE + IF,
                                          outcome = "DID",
                                          exposure = "OP",
                                          latent = c("EB", "HL", "SM", "OSE"),  # used to indicate social environment
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              RE = 4, IF = 2,
                                                              EB = 1, HL = 2.5,  SM = 3, OSE = 4,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              RE = 4, IF = 3,
                                                              EB = -1, HL = -2,  SM = -5, OSE = -8,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_SC) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# EB = Economic background
# HL = Homelessness
# SM = Sexual Minority Status
# OSE = Other Forms of Social Exclusion

```


## Risk environment

"The risk environment represents a number of structural components that moderate the relationship between individual factors and injection drug users’ (IDUs’) health. For example, higher levels of drug law enforcement (DLE) have been linked to increased risk of overdose deaths [16]. The hypothesized agent of action is that increased criminal justice risk (CJR) associated with drug use leads to specific behaviors, e.g. not calling for emergency help (i.e. calling 911 in the United States) following a witnessed overdose, which increase individual harm associated with drug use [17]. Other structural factors such as needle exchange (NE), naloxone distribution (ND), safe injection rooms (SIR) and the availability of methadone treatment (MT) can reduce the risk of overdose or harm associated with overdose in areas where these services are available"
(The relationship between US heroin market dynamics and heroin-related overdose, 1992–2008)

```{r}
assumed_relationship_RE <- dagify(DID ~ HOD + SE + NE + SIR + ND,
                                          NFNM ~ HOD + SE + SIR + JCR + ND,
                                          HOD ~ SE + IF + OP + P + DC + SIR + NE + MT,
                                          RE ~ IF + JCR,
                                          JCR ~ DLE,
                                          DLE ~ L,
                                          DC ~ OP + SE + IF, # drug components other than heroin
                                          P ~ OP + SE + IF, # purity/dose of heroin
                                          OP ~ L,
                                          latent = c("DLE", "JCR", "NE", "ND", "SIR", "MT"), # used to indicate risk environment
                                          outcome = "DID",
                                          exposure = "OP",
                                          coords = list(x = c(L = -2,
                                                              OP = 0,
                                                              SE = 6, IF = 2,
                                                              DLE = 1, JCR = 2, NE = 3, ND = 4, SIR = 5, MT = 6,
                                                              P = 2.67,
                                                              DC = 5.34,
                                                              NFNM = 10,
                                                              HOD = 8,
                                                              DID = 10), 
                                                        y =  c(L = 0,
                                                              OP = 0,
                                                              SE = 4, IF = 3,
                                                              DLE = -1, JCR = -2, NE = -3, ND = -4, SIR = -5, MT = -6,
                                                              P = 1,
                                                              DC = 1,
                                                              NFNM = 1,
                                                              HOD = 0,
                                                              DID = 0)))

ggdag_status(assumed_relationship_RE) +
  theme_dag() +
  guides(color = "none")

# L = Legality
# OP = Opioid Provision
# RE = Risk Environment
# SE = Social Environment
# IF = Individual Factors
# P = Purity
# DC = Drug Components
# NFNM = Non-Fatal Near Misses
# HOD = Heroin Overdose
# DID = Drug Induced Death 
# DLE = Drug Law Enforcement 
# CJR = Criminal justice risk
# NE = Needle Exchange
# ND = Naloxone Distribution
# MT = Methadone Treatment 
# SIR = Safe Injection Rooms

```

## PLease ignore below 

```{r}
# PLEASE IGNORE!!
commonly_assumed_relationship <- dagify(DID ~ HOD + CJR, # drug induced death
                                  HOD ~ P + DC + PDU + LOE + I + IHU + R +
                                        EB + HL + SM + OSE + SIR, # heroin-related overdose
                                  DC ~ OP, # drug components other than heroin
                                  P ~ OP + ND + NE, # purity/dose of heroin
                                  NFNM ~ HOD, # non-fatal near misses
                                  OP ~ L, # legalization and opioid provision
                                  CJR ~ DLE,
                        outcome = "DID",
                        # latent = c("P", "OP"),
                        exposure = "OP",
                        coords = list(x = c(L = 0,
                                            DID = 15, HOD = 13,
                                            OP = 1,  P = 2, DC = 2, NFNM = 15,
                                            IHU = 2, R = 3, PDU = 4, LOE = 5, I = 6, 
                                            EB = 2, HL = 3,  SM = 4, OSE = 5,
                                            DLE = 5, JCR = 6, NE = 7, ND = 8, SIR = 9, MT = 4),
                                      y = c(OP = 1, L = 1, DID = 1, HOD = 1, P = 1.5, DC = 0.5, NFNM = 1.5, PDU = 2, LOE = 2, I = 2, IHU = 2, R = 2,
                                            EB = 0, HL = 0,  SM = 0, OSE = 0,
                                            DLE = 1.5, JCR = 1.5, NE = 1.5, ND = 1.5, SIR = 1.5, MT = 1.5)))

ggdag_status(commonly_assumed_relationship) +
  theme_dag() +
  guides(color = "none")  # Turn off legend

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
